Mission-time LTL (MLTL) Formula Validation Via Regular Expressions\thanks{Work supported in part by NSF grant DMS-1950583}

### 2023-06-15 paper/TACASPaper.tex
updated paper

%KYR: need to ensure included packages don't interfere with LNCS style
%\usepackage[utf8]{inputenc}
%\usepackage{amsthm, amsmath, amssymb}
\usepackage{amsfonts} %mathbb
\usepackage{amsmath} %hdots
\usepackage{amssymb} %blacksquare
\usepackage{mathabx} %vDash
\usepackage{mathrsfs} %mathscr
\usepackage{blkarray} %for blockarray
%\usepackage{textcomp}
%\usepackage{xcolor}
%\usepackage{soul}
%\usepackage{ dsfont }
\usepackage{hyperref}
\usepackage{cancel} %for \cancel
\usepackage{algorithm} %for algorithm environment
\usepackage{algcompatible}
\usepackage[noend]{algpseudocode}
%\usepackage{changepage}%http://ctan.org/pkg/changepage
\usepackage{graphicx} %for includegraphics
\usepackage{tikz}
\usepackage{lscape} %for landscape orientation
%\usepackage{float} %for figures
\usetikzlibrary{shapes.misc, positioning}
\usetikzlibrary{arrows.meta}
\tikzset{>={Latex[width=2mm,length=2mm]}}
\usepackage{caption}
\usepackage{mwe}
%% \hypersetup{
%%     colorlinks=true,
%%     linkcolor=blue,
%%     filecolor=magenta,
%%     urlcolor=red,
%%     citecolor=teal
%% }

%LNCS:
% If you use the hyperref package, please uncomment the following line
% to display URLs in blue roman font according to Springer's eBook style:
\renewcommand\UrlFont{\color{blue}\rmfamily}

%for qed:
%% \makeatletter
%% \newcommand{\pushright}[1]{\ifmeasuring@#1\else\omit\hfill$\displaystyle#1$\fi\ignorespaces}
%% \newcommand{\pushleft}[1]{\ifmeasuring@#1\else\omit$\displaystyle#1$\hfill\fi\ignorespaces}
%% \makeatother
%% \makeatletter
%% \newcommand{\specialcell}[1]{\ifmeasuring@#1\else\omit$\displaystyle#1$\ignorespaces\fi}
%% \makeatother

@@

@@

\renewcommand{\topfraction}{.99} %figures can take up at most 99% of the page before being alone
\renewcommand{\bottomfraction}{.99} %figures can take up at most 99% of the page before being alone
\renewcommand{\textfraction}{.01} %at most this % of page will be text before making figure-only page
\addtolength{\textfloatsep}{-5mm}
\addtolength{\floatsep}{-5mm}
\addtolength{\intextsep}{-5mm}
\addtolength{\abovecaptionskip}{-3mm}
\addtolength{\belowcaptionskip}{-2mm}

@@

\begin{document}
%
%\title{Contribution Title\thanks{Supported by organization x.}}
\title{Mission-time LTL (MLTL) Formula Validation Via Regular Expressions\thanks{Work supported in part by [omitted for blind review] {\bf Note that all references to appendices will be replaced with URLs in the final paper; they are appendices only for blind review.}}}
%
\titlerunning{MLTL Formula Validation Via Regular Expressions}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\vspace{-0.1in}
%\author{First Author\inst{1}\orcidID{0000-1111-2222-3333} \and
%Second Author\inst{2,3}\orcidID{1111-2222-3333-4444} \and
%Third Author\inst{3}\orcidID{2222--3333-4444-5555}}
%\author{Jenna Elwing\thanks{University of Michigan, Ann Arbor} \and Laura Gamboa Guzmán\thanks{Iowa State University} \and Jeremy Sorkin\thanks{The Ohio State University} \and Chiara Travesset\thanks{Purdue University} \and Zili Wang\thanks{University of California, Berkeley} \and Kristin Y. Rozier\footnotemark[2]}
%
%\authorrunning{F. Author et al.}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
%\institute{Princeton University, Princeton NJ 08544, USA \and
%Springer Heidelberg, Tiergartenstr. 17, 69121 Heidelberg, Germany
%\email{lncs@springer.com}\\
%\url{http://www.springer.com/gp/computer-science/lncs} \and
%ABC Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
%\email{\{abc,lncs\}@uni-heidelberg.de}}
%
\maketitle              % typeset the header of the contribution
%

\vspace{-0.3in}
\begin{abstract}
  \noindent Mission-time Linear Temporal Logic (MLTL) represents the most practical fragment of Metric Temporal Logic; MLTL resembles the popular logic LTL with finite closed-interval integer bounds on the temporal operators. Increasingly, many tools reason over MLTL specifications, yet these tools are only useful when system designers can validate the input specifications. We design an automated characterization of the structure of the computations that satisfy a given MLTL formula using regular expressions. We prove soundness and completeness of our structure. We also give an algorithm for automated MLTL formula validation and analyze its complexity both theoretically and experimentally. Additionally, we generate a test suite using control flow diagrams to robustly test our implementation and release an open-source tool. The result of our contributions are improvements to existing algorithms for MLTL analysis applicable to many other tools and the WEST tool for automated, efficient MLTL formula validation.
  %
%        Our work connects existing tools to enable improvmeents in satisfiabiity tools and automated synthesis of verified code from MLTL behavior descriptions.
% 
  %
%  A computation is a finite sequence of truth assignments. In this paper, we use regular expressions to describe the structure of the computations that satisfy a given MLTL formula. We prove soundness and completeness of our structure. We also give an implemented algorithm (the WEST program) and analyze its complexity both theoretically and experimentally. Finally, we generate a test suite using control flow diagrams to robustly test the code.

@@

        \vspace{-0.1in}
\keywords{Mission-time Linear Temporal Logic (MLTL)  \and MLTL Satisfiability \and Temporal Logic Validation.}
\end{abstract}

\vspace{-0.4in}
\section{Introduction}
\vspace{-0.1in}
System specifications, such as aerospace operational concepts, often utilize timelines to express critical requirements. We can cite examples of this from NASA's Automated Airspace Concept \cite{EH10}, the U.~S.~Navy's Aircraft Carrier Deck Scheduler \cite{RCRBS11}, the JAXA-NASA Global Precipitation Measurement (GPM) Observatory \cite{GPM}, and many others. Formal methods provide continuously advancing tools and techniques to rigorously analyze timelines expressed in the form of temporal logic requirements, from early design-time model checking and theorem proving to on-board runtime verification. The U.~S.~Federal Aviation Administration (FAA) even advocates the use of formal methods for flight certification of these critical systems \cite{DO-178C,DO-254,DO-333}. Yet, a significant hurdle to the use of formal methods remains: how to convincingly demonstrate to the humans in the loop, from system designers to certifiers, that the analyzed formulas truly represent the system requirements \cite{Roz16}. We creatively address this validation question using regular expressions and truth tables.

@@

There are a number of SAT solvers for MLTL \textcolor{red}{CITE SAT SOLVERS IDK THEM}, but there are currently no tools to check for MLTL formula validation, i.e. there are no tools available that allow people to \emph{validate} that a proposed MLTL formula is indeed capturing a specification for a system they want to verify. %It is necessary to perform validation before even being able to utilize SAT solvers in any meaningful way. This is precisely the role of the WEST tool, which is introduced in this paper.
In this paper we introduce the WEST tool, which takes an MLTL formula and produces a description of the set of all finite time lines (of a fixed length) that satisfy the given formula, similar to filling out a truth table for propositional formulas and looking at the rows on which the formula holds. As a consequence, MLTL validation can be done by verifying that the output of the WEST program on a given input indeed matches the behaviour of the specification in question.

We show that our contributions not only fill a critical gap in temporal logic validation but also directly connect to parallel developments to enable better temporal logic formula analysis, benchmark generation, proof generation (e.g., in ACL2), and synthesis of verified C code from temporal logic behavior descriptions.

@@

We structure the paper around our contributions as follows. In Section \ref{prereq}, we build on the semantics of MLTL to define a computation and its bit string representation. % of a computation that is used throughout the paper and the WEST program. Finally, we describe negation normal form (NNF), a standardized way to express MLTL formulas. By writing the MLTL formulas in NNF, the WEST program is able to run much more efficiently.
%
In Section \ref{regex}, we recursively define regular expressions encapsulating the satisfying computations of MLTL formulas and prove that our new translation algorithm is both sound and complete. In addition, we provide a calculation for the minimum computation length required to describe all the satisfying computations of a MLTL formula that slightly improves upon existing calculations in the literature. Finally, we show an application of the regular expressions of the satisfying computations by using them to prove an MLTL theorem. In particular, we show that $\phi\mathcal{U}_{[a,b]}(\phi\mathcal{U}_{[0,c]}\psi) \equiv \phi\mathcal{U}_{[a,b+c]}\psi$ and $\phi\mathcal{R}_{[a,b]}(\phi\mathcal{R}_{[0,c]}\psi) \equiv \phi\mathcal{R}_{[a,b+c]}\psi$.
%
We introduce a new tool, called WEST (acronym redacted for blind review), that implements automated validation in Section \ref{WESTalg}
and calculate its space and time complexity, both theoretically and experimentally. We show that the theoretical worst-case complexity for both space and time is doubly exponential, but that the average case fares far better.
%
Section \ref{test} proves correctness of the algorithms and provides a test suite to show correctness of implementation with high confidence. Intelligent fuzzing techniques contribute to test suite construction from a state diagram representing the control flow of the algorithm. We also verify the correctness of outputs of the WEST program against a na\"ive brute force implementation.
%
Section \ref{simpsection}, provides a combinatorial theorem for simplifying certain outputs of the WEST program. Formulas of the form $\phi \equiv \psi$ to check for equivalence between two expressions can lead to outputs that are non-trivial to simplify. Section \ref{Conclusion} discusses impacts and future work.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vspace{-0.1in}
\section{Preliminaries: Mission-time LTL and Bit String Computations} \label{prereq}
\vspace{-0.1in}
%\subsection{Mission-time Linear Temporal Logic}
\emph{Mission-time Linear Temporal Logic (MLTL)} \cite{LVR22} is a finite variation of LTL over bounded, closed, discrete intervals of the form $[a,b]$ where $a,b \in \mathbb{N}$ and $0 \leq a \leq b$. The syntax of MLTL formulas, $\phi$ and $\psi$, over a set of atomic propositions $\mathcal{AP}$, where $p \in \mathcal{AP}$ is a propositional variable, is given by the following BNF grammar:%
\vspace{-0.1in}
$$\phi, \psi := \top \ | \ \bot \ | \ p \ | \ \neg \phi \ | \ \phi \land \psi \ | \ \phi \lor \psi \ | \ \mathcal{F}_{[a,b]} \phi \ | \ \mathcal{G}_{[a,b]} \phi \ | \ \phi \mathcal{U}_{[a,b]} \psi \ | \ \phi \mathcal{R}_{[a,b]} \psi \footnote{For simplicity, we do not include parentheses in the grammar, but the WEST program requires parentheses (see Section \ref{WESTalg}). We encode Release directly rather than as the dual of Until \cite{bmc,LVR22}.}$$
MLTL formulas can be interpreted using both finite and infinite ``time lines'' that are called \emph{computations}, which represent a discrete sequence of time instances and the truth values for the propositional variables on each one of these. For the purpose of this paper, we are only going to deal with \emph{finite} computations, that is, computations that represent only finitely many time steps. 
% \vspace{-0.1in}
%
\begin{definition}[\emph{Finite computations}]
A \emph{computation} $\pi$ of length $m$ is a sequence $\{\pi[i]\}_{i = 0}^{m-1}$ of sets of propositional variables, $\pi[i] \subseteq \mathcal{AP}$, where the $i^{th}$ set holds the propositional variables that are satisfied at the $i^{th}$ time step. That is, a propositional variable $p$ is true at time step $i$ if and only if $p \in \pi[i]$. {\color{blue} What is the semantics for the other logical and temporal operators? Where can we find a formal definition for this? This is never explicitly said before \texttt{reg} is introduced and it can lead to misinterpretations from the reviewers.} We denote the suffix of $\pi$ starting at $i$ (including $i$) by $\pi_i$. Note that $\pi_0 = \pi$.
\end{definition}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vspace{-0.1in}
\section{MLTL Regular Expressions} \label{regex}
\vspace{-0.1in}
We modify the standard definition of Regular Expressions (Regex) to introduce notation that describes the satisfying computations of an MLTL formula. We begin by quoting the parts of the standard definition of a regex from \cite{sipser} that we use to describe our computations:

@@

\begin{definition}[\emph{Temporal Regular Expression}] \label{TRE} 
Let $R$ and $T$ denote regular expressions, and let $S$ be an abbreviation for $(0 \ \lor \ 1)$.\\
Let fixed $n \in \mathbb{N}$ denote the number of propositional variables in a MLTL formula. We use the following operations to describe the form of satisfying computations of the formula in the bit string representation:
\begin{itemize}
\item Pad($R$, $T$) determines which regular expression is longer and concatenates $(,S^n)$ repeatedly to the end of the shorter regular expression until the two regular expressions are the same length. Note that in the bit string representation, $(,S^n)$ denotes a time step in which the truth value of all $n$ propositional variables does not matter.
\item $R \land T$ is the intersection of the sets of strings described by $R$ and $T$. To perform this operation, we first use Pad($R$, $T$), and then take the set intersection of the sets of strings described by the two regular expressions.
\item $R^i$ denotes regular expression consisting of $R$ repeated $i$ times for $i \geq 0$. $R^0 = \epsilon$. %KYR: is this one relevant or should we take it out?
\end{itemize}
\end{definition}
Note that our regular expressions do not use the Kleene star. This is because are computations are of a fixed, finite length.
%An example of these operations follows.
\begin{example} %KYR: NICE example!
Let $n = 2$, and let $R = S1$ and $T = 1S,1S \lor S1,S1$.
To compute $R \land T$ and $R \lor T$, we use Pad($R$,$T$). Since $T$ is the longer regex by one time step, we concatenate $(,S^2)$ to the right of $R$. Thus $T =1S,1S \lor S1,S1$ and $R = S1,SS$. Now the regexes are the same length, so we can perform set intersection and alternation:
\begin{itemize}
    \item $R \land T = 11,1S \lor S1, S1$.
    \item $R \lor T = S1,SS \lor 1S,1S \lor S1,S1 = S1,SS \lor 1S,1S$.
\end{itemize}
\end{example}

To describe the bit string representations of the satisfying computations for a given MLTL formula, we use the alphabet $\Sigma =$ \{``$0$'' , ``$1$'', ``$,$''\} and define $S$ as an abbreviation for $(0 \ \lor \ 1)$.
Let $p_0, p_1, ..., p_{n-1}$ be propositional variables and $\phi, \psi$ well-formed MLTL formulas in negation normal form. We define the regular expression of satisfying computations for a MLTL formula as follows:\\

% ZILI: Left Aligned Version
\vspace{-7mm}
\hspace{-8.5mm}
\begin{minipage}{0.45 \textwidth}
    \begin{align}
    &\text{reg}(\top) = S^n\\
    &\text{reg}(p_k) = S^{k}1S^{n-k-1} \\
    &\text{reg}(\phi \lor \psi) =  \text{reg}(\phi) \lor \text{reg}(\psi)\\
    &\text{reg}(\mathcal{G}_{[a,b]} \phi) = \bigwedge_{i=a}^{b} (S^n,)^i \text{reg}(\phi)
    \end{align}
\end{minipage}
\begin{minipage}{0.55\textwidth}
    \begin{align}
    &\text{reg}(\bot) = \emptyset \\
    &\text{reg}(\neg p_k) = S^{k}0S^{n-k-1} \\
    &\text{reg}(\phi \land \psi) = \text{reg}(\phi) \land \text{reg}(\psi) \\
    &\text{reg}(\mathcal{F}_{[a,b]} \phi) = \bigvee_{i=a}^{b} (S^n,)^i \text{reg}(\phi)
    \end{align}
\end{minipage}
\vspace{-5mm}\\
\begin{minipage}{0.99\textwidth}
\begin{align}
&\text{reg}(\phi \ \mathcal{U}_{[a,b]} \psi) =  \bigvee_{i=a}^{b} \text{reg}\left(\mathcal{G}_{[a,i-1]}\phi \land \mathcal{G}_{[i, i]} \psi\right)\\
\vspace*{-10mm}&\text{reg}(\phi \mathcal{R}_{[a,b]} \psi) =  \text{reg}\left(\mathcal{G}_{[a,b]}\psi\right) \lor \bigvee_{i=a}^{b-1} \text{reg}\left(\mathcal{G}_{[a,i]}\psi \land \mathcal{G}_{[i, i]} \phi\right) \label{release}
\end{align}
\end{minipage}
\begin{definition}[\emph{Computation Length}]
We define the computation length $\emph{cplen}(\phi)$ of an MLTL formula $\phi$ recursively:\\
          %\begin{align}
\indent \qquad   $\emph{cplen}(p_k) = \emph{cplen}(\neg p_k) = 1$\\

>>>>

%KYR: need to ensure included packages don't interfere with LNCS style
%\usepackage[utf8]{inputenc}
%\usepackage{amsthm, amsmath, amssymb}
\usepackage{amsfonts} %mathbb
\usepackage{amsmath} %hdots
\usepackage{amssymb} %blacksquare
\usepackage{mathabx} %vDash
\usepackage{mathrsfs} %mathscr
\usepackage{blkarray} %for blockarray
%\usepackage{textcomp}
%\usepackage{xcolor}
%\usepackage{soul}
%\usepackage{ dsfont }
\usepackage{hyperref}
\usepackage{cancel} %for \cancel
\usepackage{algorithm} %for algorithm environment
\usepackage{algcompatible}
\usepackage[noend]{algpseudocode}
%\usepackage{changepage}%http://ctan.org/pkg/changepage
\usepackage{graphicx} %for includegraphics
\usepackage{tikz}
\usepackage{lscape} %for landscape orientation
%\usepackage{float} %for figures
\usetikzlibrary{shapes.misc, positioning}
\usetikzlibrary{arrows.meta}
\tikzset{>={Latex[width=2mm,length=2mm]}}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{mwe}

% Spacing: left, before, after
\usepackage{titlesec}
\titlespacing*{\section}{0pt}{10pt}{2pt}
\titlespacing*{\subsection}{0pt}{5pt}{2pt}

%% \hypersetup{
%%     colorlinks=true,
%%     linkcolor=blue,
%%     filecolor=magenta,
%%     urlcolor=red,
%%     citecolor=teal
%% }

%LNCS:
% If you use the hyperref package, please uncomment the following line
% to display URLs in blue roman font according to Springer's eBook style:
\renewcommand\UrlFont{\color{blue}\rmfamily}
%for qed:
%% \makeatletter
%% \newcommand{\pushright}[1]{\ifmeasuring@#1\else\omit\hfill$\displaystyle#1$\fi\ignorespaces}
%% \newcommand{\pushleft}[1]{\ifmeasuring@#1\else\omit$\displaystyle#1$\hfill\fi\ignorespaces}
%% \makeatother
%% \makeatletter
%% \newcommand{\specialcell}[1]{\ifmeasuring@#1\else\omit$\displaystyle#1$\ignorespaces\fi}
%% \makeatother

@@

% Used in REST experiment section
\usepackage{wrapfig}

@@

\renewcommand{\topfraction}{.99} %figures can take up at most 99% of the page before being alone
\renewcommand{\bottomfraction}{.99} %figures can take up at most 99% of the page before being alone
\renewcommand{\textfraction}{.01} %at most this % of page will be text before making figure-only page
\addtolength{\textfloatsep}{-5mm}
\addtolength{\floatsep}{-5mm}
\addtolength{\intextsep}{-5mm}
\addtolength{\abovecaptionskip}{-3mm}
\addtolength{\belowcaptionskip}{-2mm}
\addtolength{\abovedisplayskip}{-5pt}
\addtolength{\belowdisplayskip}{-5pt}

@@

\begin{document}
%
%\title{Contribution Title\thanks{Supported by organization x.}}
\title{Mission-time LTL (MLTL) Formula Validation Via Regular Expressions\thanks{Work supported in part by NSF grant DMS-1950583}}
%
\titlerunning{MLTL Formula Validation Via Regular Expressions}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%

\author{Jenna Elwing\inst{1}\and
Laura Gamboa-Guzman\inst{2} \and
Jeremy Sorkin\inst{3} \and Chiara Travesset\inst{4}\and Zili Wang\inst{2} \and Kristin Y. Rozier\inst{2}}
%
\authorrunning{J. Elwing et al.}
\institute{University of Michigan, Ann Arbor \and
Iowa State University \and University of Amsterdam \and University of Wisconsin, Madison}

\maketitle
\vspace{-0.3in}
\begin{abstract}
  \noindent Mission-time Linear Temporal Logic (MLTL) represents the most practical fragment of Metric Temporal Logic; MLTL resembles the popular logic LTL with finite closed-interval integer bounds on the temporal operators. Increasingly, many tools reason over MLTL specifications, yet these tools are only useful when system designers can validate the input specifications. We design an automated characterization of the structure of the computations that satisfy a given MLTL formula using regular expressions. We prove soundness and completeness of our structure. We also give an algorithm for automated MLTL formula validation and analyze its complexity both theoretically and experimentally. Additionally, we generate a test suite using control flow diagrams to robustly test our implementation and release an open-source tool. The result of our contributions are improvements to existing algorithms for MLTL analysis applicable to many other tools and the WEST tool for automated, efficient MLTL formula validation.
  %
%        Our work connects existing tools to enable improvmeents in satisfiabiity tools and automated synthesis of verified code from MLTL behavior descriptions.
% 
  %
%  A computation is a finite sequence of truth assignments. In this paper, we use regular expressions to describe the structure of the computations that satisfy a given MLTL formula. We prove soundness and completeness of our structure. We also give an implemented algorithm (the WEST program) and analyze its complexity both theoretically and experimentally. Finally, we generate a test suite using control flow diagrams to robustly test the code.

@@

% \vspace{-0.1in}
\keywords{Mission-time Linear Temporal Logic (MLTL)  \and MLTL Validation \and Temporal Logic Validation.}
\end{abstract}

% \vspace{-0.4in}
\section{Introduction}
% \vspace{-0.1in}
System specifications, such as aerospace operational concepts, often utilize timelines to express critical requirements. We can cite examples of this from NASA's Automated Airspace Concept \cite{EH10}, the U.~S.~Navy's Aircraft Carrier Deck Scheduler \cite{RCRBS11}, the JAXA-NASA Global Precipitation Measurement (GPM) Observatory \cite{GPM}, and many others. Formal methods provide continuously advancing tools and techniques to rigorously analyze timelines expressed in the form of temporal logic requirements, from early design-time model checking and theorem proving to on-board runtime verification. The U.~S.~Federal Aviation Administration (FAA) even advocates the use of formal methods for flight certification of these critical systems \cite{DO-178C,DO-254,DO-333}. Yet, a significant hurdle to the use of formal methods remains: how to convincingly demonstrate to the humans in the loop, from system designers to certifiers, that the analyzed formulas truly represent the system requirements \cite{Roz16}. We creatively address this validation question using regular expressions and truth tables.

@@

There exists a SAT solver for MLTL, MLTLSAT \cite{LVR22}, but there are currently no tools for MLTL formula validation. This paper introduces the WEST tool \href{https://github.com/zwang271/2022-Iowa-State-REU-Temporal-Logic-}{[GitHub]}, which produces a description of the set of all finite time lines (of a fixed length) that satisfy a given MLTL formula, similar to a truth table for propositional formulas. MLTL validation can be done by verifying that the output of the WEST program indeed matches the behaviour of the specification in question.

%The WEST program utilizes regular expressions to describe and visualize the finite time lines satisfying an MLTL formula, an approach that differs from the more commonly used ones that rely on finite automata theory. Currently, there are no tools that visualize MLTL formulas in the form of a finite automata, but even if such a tool were to exist, it would not fill the same niche for formula validation as the WEST algorithm. For example, the WEST algorithm allows validation and visualization for subformulas of any given formula, and allows visualization of the truth value for these at each time step of the satisfying computations, while automata do not necessarily have these capabilities.

We show that our contributions not only fill a critical gap in temporal logic validation but also directly connect to parallel developments to enable better temporal logic formula analysis, benchmark generation, proof generation (e.g., in ACL2), and synthesis of verified C++ code from temporal logic behavior descriptions.

@@

We structure the paper as follows. Section \ref{prereq} builds on the semantics of MLTL to define a computation and its bit string representation. % of a computation that is used throughout the paper and the WEST program. Finally, we describe negation normal form (NNF), a standardized way to express MLTL formulas. By writing the MLTL formulas in NNF, the WEST program is able to run much more efficiently.
%
Section \ref{regex}, recursively defines regular expressions encapsulating the satisfying computations of MLTL formulas. We provide a calculation for the minimum computation length required to describe all the satisfying computations of a MLTL formula that slightly improves upon existing calculations in the literature. Finally, we show an application of the regular expressions by using them to prove an MLTL theorem.
%
We introduce the WEST tool that implements automated validation in Section \ref{WESTalg}
and calculate its space and time complexity, both theoretically and experimentally. %We show that the theoretical worst-case complexity for both space and time is doubly exponential, but that the average case fares far better.
%
Section \ref{test} proves the correctness of WEST and provides a test suite to show correctness of implementation with high confidence. Intelligent fuzzing techniques contribute to test suite construction from a state diagram representing the control flow of WEST. We also verify the correctness of outputs of the WEST program against a na\"ive brute force implementation.
%
Section \ref{simpsection} provides a combinatorial theorem for simplifying certain outputs of the WEST program to the trivial computation. %Formulas of the form $\phi \equiv \psi$ to check for equivalence between two expressions can lead to outputs that are non-trivial to simplify. 
%
Section \ref{Using WEST} demonstrates a specific use case of the WEST tool and explores the currently supported features. 
%
Section \ref{Conclusion} discusses impacts and future work. 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \vspace{-0.40in}
\section{Preliminaries: Mission-time LTL and Bit String Computations} \label{prereq}
% \vspace{-0.5in}
%\subsection{Mission-time Linear Temporal Logic}
\emph{Mission-time Linear Temporal Logic (MLTL)} \cite{LVR22} is a finite variation of LTL over bounded, closed, discrete intervals of the form $[a,b]$ where $a,b \in \mathbb{N}$ and $0 \leq a \leq b$. The syntax of MLTL formulas, $\phi$ and $\psi$, over a set of atomic propositions $\mathcal{AP}$, where $p \in \mathcal{AP}$ is a propositional variable, is given by the following BNF grammar: \\
% \vspace{-0.1in}
$$\phi, \psi := \top \ | \ \bot \ | \ p \ | \ \neg \phi \ | \ \phi \land \psi \ | \ \phi \lor \psi \ | \ \mathcal{F}_{[a,b]} \phi \ | \ \mathcal{G}_{[a,b]} \phi \ | \ \phi \mathcal{U}_{[a,b]} \psi \ | \ \phi \mathcal{R}_{[a,b]} \psi.\footnote{For simplicity, we do not include parentheses in the grammar, but the WEST program requires parentheses (see Section \ref{WESTalg}). We encode Release directly rather than as the dual of Until \cite{bmc,LVR22}.}$$

MLTL formulas can be interpreted using both finite and infinite ``time lines'' that are called \emph{computations}, which represent a discrete sequence of time instances and the truth values for the propositional variables on each one of these. For the purpose of this paper, we are only going to deal with \emph{finite} computations, that is, computations that represent only finitely many time steps. 
% \vspace{-0.1in}
%
\begin{definition}[\emph{Finite computations}]
A \emph{computation} $\pi$ of length $m$ is a sequence $\{\pi[i]\}_{i = 0}^{m-1}$ of sets of propositional variables, $\pi[i] \subseteq \mathcal{AP}$, where the $i^{th}$ set holds the propositional variables that are satisfied at the $i^{th}$ time step. That is, a propositional variable $p$ is true at time step $i$ if and only if $p \in \pi[i]$. We denote the suffix of $\pi$ starting at $i$ (including $i$) by $\pi_i$. Note that $\pi_0 = \pi$.
\end{definition}
We now provide the semantics for MLTL. A computation $\pi$ satisfies a given MLTL formula $\alpha$, written $\pi \vDash \alpha$, in the following cases\footnote{We do not include the Next operator, often denoted by $\chi$, as it can be defined as $\mathcal{G}_{[1,1]}$.}:\\
\begin{minipage}{0.5\textwidth}
\begin{flalign*}
   &\pi \vDash p \text{ iff } p \in \pi[0]&\\
   &\pi \vDash \alpha \land \beta \text{ iff } \pi \vDash \alpha \text{ and } \pi \vDash \beta&
\end{flalign*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{flalign*}
    &\pi \vDash \neg \alpha \text{ iff } \pi \nvDash \alpha&\\
    &\pi \vDash \alpha \lor \beta \text{ iff } \pi \vDash \alpha \text{ or } \pi \vDash \beta&
\end{flalign*}
\end{minipage}
{\setlength{\abovedisplayskip}{0pt}
\begin{flalign*}
    &\pi \vDash \mathcal{F}_{[a,b]}\alpha \text{ iff } |\pi| > a \text{ and } \exists i \in [a,b] \text{ such that } \pi_i \vDash \alpha&\\
    &\pi \vDash \mathcal{G}_{[a,b]}\alpha \text{ iff } |\pi| \leq a \text{ or } \forall i \in [a,b] \ \pi_i \vDash \alpha&\\
    &\pi \vDash \alpha \ \mathcal{U}_{[a,b]} \beta \text{ iff } |\pi| > a \text{ and } \exists i \in [a,b] \text{ such that } \pi_i \vDash \beta \text{ and }\forall j \in [a, i-1]\ \pi_j \vDash \alpha &\\
    &\pi \vDash \alpha \ \mathcal{R}_{[a,b]} \beta \text{ iff } |\pi| \leq a \text{ or } \forall i \in [a,b] \ \pi_i \vDash \beta \text{ or }\exists j \in [a,b-1] \text{ such that } \pi_j \vDash \alpha &\\
    &\indent \text{ and } \forall a \leq k \leq j \ \pi_k \vDash \beta&
\end{flalign*}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \vspace{-0.1in}
\section{MLTL Regular Expressions} \label{regex}
% \vspace{-0.1in}
We modify the standard definition of Regular Expressions (Regex) to introduce notation that describes the satisfying computations of an MLTL formula. We begin by quoting the parts of the standard definition of a regex from \cite{sipser} that we use to describe our computations:

@@

\begin{definition}[\emph{Temporal Regular Expression}] \label{TRE} 
Let $R$ and $T$ denote regular expressions, and let $S$ be an abbreviation for $(0 \ \lor \ 1)$.\\
Let fixed $n \in \mathbb{N}$ denote the number of propositional variables in a MLTL formula. We use the following operations to describe the form of satisfying computations of the formula in the bit string representation:
\begin{itemize}
\item Pad($R$, $T$) determines which regular expression is longer and concatenates $(,S^n)$ repeatedly to the end of the shorter regular expression until the two regular expressions are the same length. Note that in the bit string representation, $(,S^n)$ denotes a time step in which the truth value of all $n$ propositional variables does not matter.
\item $R \land T$ is the intersection of the sets of strings described by $R$ and $T$. To perform this operation, we first use Pad($R$, $T$), and then take the set intersection of the sets of strings described by the two regular expressions.
\item $R^i$ denotes regular expression consisting of $R$ repeated $i$ times for $i \geq 0$. $R^0 = \epsilon$. %KYR: is this one relevant or should we take it out?
\end{itemize}
\end{definition}
Note that our regular expressions do not use the Kleene star. This is because our computations are of a fixed, finite length.
%An example of these operations follows.
\begin{example} %KYR: NICE example!
Let $n = 2$, and let $R = S1$ and $T = 1S,1S \lor S1,S1$.
To compute $R \land T$ and $R \lor T$, we use Pad($R$,$T$). Since $T$ is the longer regex by one time step, we extend $R$ by one time-step. Thus $T =1S,1S \lor S1,S1$ and $R = S1,SS$. Now we can perform set intersection and alternation on the two regular expressions of equal length:
\begin{itemize}
    \item $R \land T = 11,1S \lor S1, S1$.
    \item $R \lor T = S1,SS \lor 1S,1S \lor S1,S1 = S1,SS \lor 1S,1S$.
\end{itemize}
\end{example}

To describe the bit string representations of the satisfying computations for a given MLTL formula, we use the alphabet $\Sigma =$ \{``$0$'' , ``$1$'', ``$,$''\} and define $S$ as an abbreviation for $(0 \ \lor \ 1)$.
Let $p_0, p_1, ..., p_{n-1}$ be propositional variables and $\phi, \psi$ well-formed MLTL formulas in negation normal form. We define the regular expression of satisfying computations for a MLTL formula as follows:\\
\begin{minipage}{0.5\textwidth}
    \begin{flalign*}
        &\text{reg}(\top) = S^n&\\
    &\text{reg}(p_k) = S^{k}1S^{n-k-1}&\\
    &\text{reg}(\phi \lor \psi) =  \text{reg}(\phi) \lor \text{reg}(\psi)&\\
    &\text{reg}(\mathcal{G}_{[a,b]} \phi) = \bigwedge_{i=a}^{b} (S^n,)^i \text{reg}(\phi)&
    \end{flalign*}
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{flalign*}
        &\text{reg}(\bot) = \emptyset&\\
    &\text{reg}(\neg p_k) = S^{k}0S^{n-k-1}&\\
    &\text{reg}(\phi \land \psi) = \text{reg}(\phi) \land \text{reg}(\psi)&\\
    &\text{reg}(\mathcal{F}_{[a,b]} \phi) = \bigvee_{i=a}^{b} (S^n,)^i \text{reg}(\phi)&
\end{flalign*}    
\end{minipage}
{\setlength{\abovedisplayskip}{0pt}
\begin{flalign*}
    &\text{reg}(\phi \ \mathcal{U}_{[a,b]} \psi) =  \bigvee_{i=a}^{b} \text{reg}\left(\mathcal{G}_{[a,i-1]}\phi \land \mathcal{G}_{[i, i]} \psi\right)&\\
&\text{reg}(\phi \mathcal{R}_{[a,b]} \psi) =  \text{reg}\left(\mathcal{G}_{[a,b]}\psi\right) \lor \bigvee_{i=a}^{b-1} \text{reg}\left(\mathcal{G}_{[a,i]}\psi \land \mathcal{G}_{[i, i]} \phi\right)&
\end{flalign*}
}
